---
title: "Using the `tvvarss` package"
author: "Eric Ward, Mark Scheuerell, Steve Katz"
date: "Apr 3, 2017"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tvvarss}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation

```{r install}
library(rstan)
library(tvvarss)
## for optimizing stan on your machine,
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

## Model Fitting

There is a single function to fit TVVARSS models: `tvvarss`. At a minimum, the user needs to specify the time series data as an array `y` where `dim(y) = c(n_sites, n_year, n_species)`.

```{r function, echo=TRUE, eval=FALSE}
tvvarss(y = y)
```

The full set of function options is

```{r function2, echo=TRUE, eval=FALSE}
tvvarss(y = y, include_trend = TRUE, de_mean = TRUE, x0 = NULL,
        shared_q = NULL, shared_r = NULL, shared_u = NULL,
        mcmc_iter = 1000, mcmc_warmup = 500, mcmc_thin = 1, mcmc_chain = 3)
```

The sampling parameters with the `mcmc_` prefix are straightforward (and need to be set to much smaller models than the same models in JAGS). However, the remaining parameters are useful for changing the model structure. 

## Data (1 site)

```{r sim_proc}
set.seed(123)
## number of time steps
TT <- 30
## number of spp/guilds
nn <- 4
## CASE 1: linear food chain
B0_lfc <- matrix(list(0),nn,nn)
for(i in 1:(nn-1)) {
  B0_lfc[i,i+1] <- "td"
  B0_lfc[i+1,i] <- "bu"
}
for(i in 1:nn) {
  B0_lfc[i,i] = "dd"
}
## simulate & plot states
lfc <- simTVVAR(B0_lfc,TT,var_QX=rev(seq(1,4)/40),cov_QX=0,var_QB=0.05,cov_QB=0)
matplot(t(lfc$states),type="l")
```

## Observation Error

We can use the `sim2fit` function to add observation error to the simulated process. It requires a `simTVVAR` object and the number of desired realizations of the process.

```{r add_obs_err}
dat_obs = sim2fit(lfc, n_sims = 1)
```

## Fitting

Now we can use `tvvarss` to estimate the parameters in the model.

```{r, eval = FALSE}
mod_fit = tvvarss(y = dat_obs, B = B, include_trend = FALSE, de_mean = TRUE, x0 = NULL,
                  shared_q = NULL, shared_r = NULL, shared_u = NULL,
                  mcmc_iter = 200, mcmc_warmup = 100, mcmc_thin = 1, mcmc_chain = 1)
```

## Validation

```{r, eval = FALSE}
## extract fitted values
pred = apply(extract(mod_fit, c("pred"))$pred, c(3,4), mean)
## plot observations and estimated states
par(mfrow = c(2,2), mgp=c(2,1,0), mai=c(0.3,0.3,0.1,0.1))
for(i in 1:4) {
  plot(pred[,i], type="l", ylim=range(c(dat_obs[,i], pred[,i])))
  points(dat_obs[,i], col="red")
}
```

## Working with parameters

The function `tvvarss` returns a `stanfit` object. This can be plotted directly, using 

```{r, eval=FALSE}
plot(mod_fit, pars = c("B[3,2,2]"))
traceplot(mod_fit, pars = c("B[3,2,2]"))
```

Or the parameters may be stored in a new object, 

```{r, eval=FALSE}
pars = extract(mod_fit, permuted = TRUE)
```

To retain the original samples by MCMC chain, set `permuted=FALSE`

Parameters may also be extracted via the `broom` package, into a data frame that just contains the estimate and standard error (where they can be passed to ggplot, etc). 

```{r, eval=FALSE}
library(broom)
tidy_pars = tidy(mod_fit)
```

## Model Selection

Fitted `tvvarss` objects can be compared to one another using the `loo` package for model selection, which can compute either the Leave One Out Information Criterion (LOOIC) or Watanabe's Akaike Information Criterion (WAIC). For example,

```{r, eval=FALSE}
library(loo)
LL <- loo(extract_log_lik(mod_fit))
## LOOIC
looic <- LL$looic
## WAIC
waic <- LL$waic
```

