---
title: "Multi-site simulations"
author: ""
date: "11/8/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simulation overview
We're generating simulated time series of a community (4 species) over 30 years (after burn-in occurs). The B matrix is the linear food chain example, 

```{r echo = FALSE}
n_species <- 4
B0_init <- matrix(0,n_species,n_species)
diag(B0_init) <- 0.6
B0_init[1,2] <- -0.03
B0_init[2,1] <- 0.1
B0_init[2,3] <- -0.1
B0_init[3,2] <- 0.1
B0_init[3,4] <- -0.3
B0_init[4,3] <- 0.1
print(B0_init)
```

The purpose of these is to look at adding replicates. There's 2 kinds of replicates: (1) repeated measurements of the same underlying state (which is what's being explored here), and (2) repeated realizations of the same community (similar to what we did with the San Nic paper).

We simulate 100 datasets using either 1, 3, or 5 sites, and then fit our Bayesian MARSS model (static - B for now) to those data to examine our ability to recover parameters. Simulated observation error is low in all these cases (0.01) as is process variance (0.02). 


## Results

We can do things like look at coverage or median estimates as a function of the number of sites. 

```{r echo = FALSE}
load("../supplement/output_linear_chain_staticB.Rdata")
d1 = saved_output
load("../supplement/output_linear_chain_staticB_multisite3.Rdata")
d3 = saved_output
load("../supplement/output_linear_chain_staticB_multisite5.Rdata")
d5 = saved_output
```

```{r fig1, echo=FALSE, fig.cap = "Posterior means of the estimates from the model with 1-site"}
load("../supplement/output_linear_chain_staticB.Rdata")
d = saved_output

par(mfrow=c(4,4), mgp=c(2,1,0), mai=c(0.1,0.01,0.1,0.01))
for(i in 1:4) {
  for(j in 1:4) {
    idx = which(startsWith(paste0("B[",1,",",i,",",j,"]"), d[[1]]$estimate$term))
    z = 0
    for(k in 1:100) {
      z[k] = d[[k]]$estimate$estimate[idx]
    }
    if(var(z) == 0) {
      plot(0,0,col="white",axes=F,main=paste0("B[",i,",",j,"]"))
    } else {
    hist(z, col="grey70", axes=F, main=paste0("B[",i,",",j,"]"))
    lines(rep(B0_init[i,j], 2), c(0, 10000), lwd=3, col="red")
    }
  }
}
```


```{r fig2, echo=FALSE, fig.cap = "Posterior means of the estimates from the model with 3-site"}
load("../supplement/output_linear_chain_staticB_multisite3.Rdata")
d = saved_output

par(mfrow=c(4,4), mgp=c(2,1,0), mai=c(0.1,0.01,0.1,0.01))
for(i in 1:4) {
  for(j in 1:4) {
    idx = which(startsWith(paste0("B[",1,",",i,",",j,"]"), d[[1]]$estimate$term))
    z = 0
    for(k in 1:100) {
      z[k] = d[[k]]$estimate$estimate[idx]
    }
    if(var(z) == 0) {
      plot(0,0,col="white",axes=F,main=paste0("B[",i,",",j,"]"))
    } else {
    hist(z, col="grey70", axes=F, main=paste0("B[",i,",",j,"]"))
    lines(rep(B0_init[i,j], 2), c(0, 10000), lwd=3, col="red")
    }
  }
}
```


```{r fig3, echo=FALSE, fig.cap = "Posterior means of the estimates from the model with 5-site"}
load("../supplement/output_linear_chain_staticB_multisite5.Rdata")
d = saved_output

par(mfrow=c(4,4), mgp=c(2,1,0), mai=c(0.1,0.01,0.1,0.01))
for(i in 1:4) {
  for(j in 1:4) {
    idx = which(startsWith(paste0("B[",1,",",i,",",j,"]"), d[[1]]$estimate$term))
    z = 0
    for(k in 1:100) {
      z[k] = d[[k]]$estimate$estimate[idx]
    }
    if(var(z) == 0) {
      plot(0,0,col="white",axes=F,main=paste0("B[",i,",",j,"]"))
    } else {
    hist(z, col="grey70", axes=F, main=paste0("B[",i,",",j,"]"))
    lines(rep(B0_init[i,j], 2), c(0, 10000), lwd=3, col="red")
    }
  }
}
```


```{r fig4, echo=FALSE, fig.cap = "Posterior coverage of the estimates from the model with 1-site"}
# coverage
load("../supplement/output_linear_chain_staticB.Rdata")
par(mfrow=c(4,4), mgp=c(2,1,0), mai=c(0.1,0.01,0.1,0.01))
for(i in 1:4) {
  for(j in 1:4) {
    idx = which(startsWith(paste0("B[",1,",",i,",",j,"]"), saved_output[[1]]$estimate$term))
    z = 0
    for(k in 1:100) {
      # log-score is dnorm(x = B0_init[i,j], mean = saved_output[[k]]$estimate$estimate[idx], sd = )
      mu = saved_output[[k]]$estimate$estimate[idx]
      sigma = saved_output[[k]]$estimate$std.error[idx]
      z[k] = ifelse(B0_init[i,j] < (mu+2*sigma) & B0_init[i,j] > (mu-2*sigma), 1, 0)
    }
    if(is.na(var(z)) || var(z) == 0) {
      plot(0,0,col="white",axes=F,main=paste0("B[",i,",",j,"]"))
    } else {
      hist(z, col="grey70", axes=F, main=paste0("B[",i,",",j,"]"))
      #lines(rep(B0_init[i,j], 2), c(0, 10000), lwd=3, col="red")
    }
  }
}
```


```{r fig5, echo=FALSE, fig.cap = "Posterior coverage of the estimates from the model with 3-site"}
# coverage
load("../supplement/output_linear_chain_staticB_multisite3.Rdata")
par(mfrow=c(4,4), mgp=c(2,1,0), mai=c(0.1,0.01,0.1,0.01))
for(i in 1:4) {
  for(j in 1:4) {
    idx = which(startsWith(paste0("B[",1,",",i,",",j,"]"), saved_output[[1]]$estimate$term))
    z = 0
    for(k in 1:100) {
      # log-score is dnorm(x = B0_init[i,j], mean = saved_output[[k]]$estimate$estimate[idx], sd = )
      mu = saved_output[[k]]$estimate$estimate[idx]
      sigma = saved_output[[k]]$estimate$std.error[idx]
      z[k] = ifelse(B0_init[i,j] < (mu+2*sigma) & B0_init[i,j] > (mu-2*sigma), 1, 0)
    }
    if(is.na(var(z)) || var(z) == 0) {
      plot(0,0,col="white",axes=F,main=paste0("B[",i,",",j,"]"))
    } else {
      hist(z, col="grey70", axes=F, main=paste0("B[",i,",",j,"]"))
      #lines(rep(B0_init[i,j], 2), c(0, 10000), lwd=3, col="red")
    }
  }
}
```

```{r fig6, echo=FALSE, fig.cap = "Posterior coverage of the estimates from the model with 5-site"}
# coverage
load("../supplement/output_linear_chain_staticB_multisite5.Rdata")
par(mfrow=c(4,4), mgp=c(2,1,0), mai=c(0.1,0.01,0.1,0.01))
for(i in 1:4) {
  for(j in 1:4) {
    idx = which(startsWith(paste0("B[",1,",",i,",",j,"]"), saved_output[[1]]$estimate$term))
    z = 0
    for(k in 1:100) {
      # log-score is dnorm(x = B0_init[i,j], mean = saved_output[[k]]$estimate$estimate[idx], sd = )
      mu = saved_output[[k]]$estimate$estimate[idx]
      sigma = saved_output[[k]]$estimate$std.error[idx]
      z[k] = ifelse(B0_init[i,j] < (mu+2*sigma) & B0_init[i,j] > (mu-2*sigma), 1, 0)
    }
    if(is.na(var(z)) || var(z) == 0) {
      plot(0,0,col="white",axes=F,main=paste0("B[",i,",",j,"]"))
    } else {
      hist(z, col="grey70", axes=F, main=paste0("B[",i,",",j,"]"))
      #lines(rep(B0_init[i,j], 2), c(0, 10000), lwd=3, col="red")
    }
  }
}
```
