@@ -1,141 +0,0 @@
---
title: "MARSS vs TVVARSS comparison"
output:
  md_document:
    variant: markdown_github
    toc: yes
    toc_depth: 3
---

### Version

This is version `r paste0('0.',format(Sys.time(), '%y.%m.%d'))`.

***

# Inits

```{r load_pkgs}
library(MARSS)
library(tvvarss)
library(MASS)
```

# MARSS model

Here is a simple predator-prey model in MAR(1) form where the predator is under relatively strong density dependence, but the prey is expected to have relatively density-independent dynamics in the absence of predation.

```{r MARSS_model}
## number of species/guilds
n_species <- 2
## number of years to simulate
n_year <- 60
## observation variance
sd_obs <- sqrt(0.05)
## initial B
BB <- matrix(0,n_species,n_species)
BB[1,1] <- 0.7
BB[1,2] <- -0.2
BB[2,1] <- 0.1
BB[2,2] <- 0.3
BB
```

Now we can simulate a realization of the process using `tvvarss::simTVVAR()`.

```{r sim_data}
simdat <- simTVVAR(Bt = BB, topo = NULL, TT = n_year,
                   var_QX = c(0.3, 0.1), cov_QX = 0,
                   QQ_BB = matrix(0, n_species*n_species, n_species*n_species))
## other method
xx <- matrix(0, n_species, n_year)
QQ <- diag(c(0.3, 0.1))
for(t in 2:n_year) {
  xx[,t] <- BB %*% xx[,t-1] + mvrnorm(1, matrix(0, n_species, 1), QQ)
}
yy <- xx + rnorm(n_species*n_year, 0, sqrt(0.1))
```

## Test I: 1 observation of process

For the first test, we will use only one observation for each of the states, and set the observation variance at 0.1. We will also throw away the first 10 data points to eliminate the effect of the initial conditions.

```{r add_obs_err}
y1 <- sim2fit(simdat, 1, sd=sd_obs, new_real = FALSE)
z1 <- t(scale(y1[,,], scale = FALSE))[,-c(1:10)]
```

Here is the model definition and fit for `MARSS()`.

```{r fit_MARSS_model, cache=TRUE}
## model list
mod_list <- list(
  B="unconstrained",
  U="zero",
  C="zero",
  c="zero",
  Q="diagonal and unequal",
  Z="identity",
  A="zero",
  D="zero",
  d="zero",
  R="diagonal and equal"
)
## control list
con_list <- list(maxit=9999, conv.test.slope.tol=0.1)
## fit MARSS
mf1 <- MARSS(y = z1, model = mod_list, control = con_list)
round(coef(mf1, type = "matrix")$B, 2)
BB
```

## Test II: 2 observations of process

For the second test, we will use 2 observations for each of the states, and set the observation variance at 0.1. Again, we'll drop the first 10 observations.

```{r add_obs_err_2}
y2 <- sim2fit(simdat, 2, sd=sd_obs, new_real = FALSE)
z2 <- t(scale(t(rbind(y2[,,1],y2[,,2])), scale = FALSE))[,-c(1:10)]
```

Here is the model definition and fit for `MARSS()`.

```{r fit_MARSS_model_2, cache=TRUE}
## new Z
ZZ <- matrix(0, n_species*2, n_species)
ZZ[1:2,1] <- 1
ZZ[3:4,2] <- 1
## model list
mod_list$Z <- ZZ
## fit MARSS
mf2 <- MARSS(y = z2, model = mod_list, control = con_list)
round(coef(mf2, type = "matrix")$B, 2)
BB
```

## Test III: 3 observations of process

For the second test, we will use 2 observations for each of the states, and set the observation variance at 0.1. Again, we'll drop the first 10 observations.

```{r add_obs_err_3}
y3 <- sim2fit(simdat, 3, sd=sd_obs, new_real = FALSE)
z3 <- t(scale(t(rbind(y3[,,1],y3[,,2])), scale = FALSE))[,-c(1:10)]
```

Here is the model definition and fit for `MARSS()`.

```{r fit_MARSS_model_3, cache=TRUE}
## new Z
ZZ <- matrix(0, n_species*3, n_species)
ZZ[1:3,1] <- 1
ZZ[4:6,2] <- 1
## model list
mod_list$Z <- ZZ
## fit MARSS
mf3 <- MARSS(y = z3, model = mod_list, control = con_list)
round(coef(mf3, type = "matrix")$B, 2)
BB
```



